---
title: "Sensory and Motor Variance in Adaptation and Proprioceptive Recalibration"
author: "Marius 't Hart"
date: "20/11/2022"
output: html_document
---

```{r setup_knitr, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Variance in movements has sometimes been seen as a measure of exploration, which is thought to support motor adaptation. For example, people with larger movement variability may adapt to a perturbation quicker.

Similarly, if localization of an unseen hand after an outward reach depends on proprioception, and this sensory signal may be recalibrated, then a maximum likelihood approach would predict that people with less reliable (i.e. more variable) proprioception, recalibrated proprioception to a larger degree.

We have published four papers on related topics, that all use the same paradigm. Here we combine all this data to have a suitable data set to probe questions related to the role that variance has in adaptation processes. For some questions, we can include pilot data where no adaptation was recorded (i.e. only aligned sessions are available).

We start by installing a package with generic functions for analyzing reach data.

```{r get_reach_package}
if (!require(Reach)) {
  if (!require(devtools)) {
    install.packages(devtools)
  }
  devtools::install_github('thartbm/Reach', ref='main', force=TRUE)
}
library(Reach)
```

Now we load all the functions used specifically for this project. They are separated in source files, grouped by topic:

```{r source_R_code}
# here are some general utility functions:
source('R/util.R')
# this does data pre-processing:
source('R/data.R')
```


Now we are ready to download the full data set:

```{r download_data}
# from the data set, we are interested in groups that all learned the
# same 30 degree rotation, without instructions


# 'control', 'instructed', 'control60', 'instructed60',
# 'cursorjump', 'handview', 'older_control', 'older_instructed'
# 'EDS', 'EDSmatch', 'org_control', 'org_instructed', 'org_control60', and
# 'org_instructed60'

groups<- c('control',
           'instructed',
           'control60',
           'instructed60',
           'cursorjump',
           'handview',
           'older_control',
           'older_instructed',
           'EDS',
           'EDSmatch'
)

downloadData(folder='data', 
             unzip=TRUE, 
             removezips=TRUE, 
             checkfiles=TRUE, 
             groups=groups, 
             sections=c('aligned','rotated'), 
             overwrite=TRUE)


groups <- c(
           'org_control',
           'org_instructed',
           'org_control60',
           'org_instructed60'
           )

downloadData(folder='data', 
             unzip=TRUE, 
             removezips=TRUE, 
             checkfiles=TRUE, 
             groups=groups, 
             sections=c('aligned','rotated'), 
             overwrite=TRUE)


```

And we can test if all of the data is there:

```{r check_data}

groups <- c('control',
            'instructed',
            'control60',
            'instructed60',
            'cursorjump',
            'handview',
            'older_control',
            'older_instructed',
            'EDS',
            'EDSmatch'
            )

sessions <- c('aligned',
              'rotated')

tasks <- c('training',
           'activelocalization',
           'passivelocalization',
           'nocursor')

dataChecks( groups   = groups, 
            sessions = sessions, 
            tasks    = tasks )
  
groups <- c('org_control',
            'org_instructed',
            'org_control60',
            'org_instructed60')
sessions <- c('aligned')

dataChecks( groups   = groups, 
            sessions = sessions, 
            tasks    = tasks )
  
```

One participants seems to be missing a block of 9 trials from each of the 4 kinds of localization, but other than that, there are only one or two trials missing from each participant. This is workable data.

As a final preparation we get a file with descriptors for several measurements for every participant. This also takes some time, but has to be done only once.

```{r get_descriptors_file}
getAllDescriptors()
```

